---
title: "ESPRI Thrive Particpant Report"
date: "`r paste('As of', Sys.Date())`"
output: 
  word_document:
    reference_docx: do-not-delete.docx
knit: (
  function(input_file, encoding) {
    output_file <- paste0(stringr::str_replace(input_file, ".Rmd", ""), "-", Sys.Date(), '.docx')
    
    rmarkdown::render( 
      input       = input_file, 
      encoding    = encoding,     
      output_file = output_file) })
---

```{r, echo=FALSE, cache=FALSE, message=FALSE}
library(knitr)
library(rmdformats)
library(dplyr)
library(readxl)
library(zoo)
library(tidyr)
library(ggplot2)
library(pander)
library(scales)
## Global options
#options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
#opts_knit$set(width=75)
```

```{r read_data}
get_newest <- function(path, pattern){
  newest <- NA
  time <- as.POSIXct("2019-01-01T00:00:00.500-08:00" ,format="%Y-%m-%dT%H:%M:%OS")
  for (f in list.files(path = path, pattern = pattern)){
    full_path <- paste0(path, f)
    ctime <- file.info(full_path)$ctime
    if (ctime > time) {
      time <- ctime
      newest <- full_path
    }
  }
  return(newest)
}

quick_tally <- function(df, var, header){
  df$var <- df[[var]]
  total <- nrow(df)
  df <- df %>% 
    group_by(var) %>% 
    tally() %>%
    mutate(total = total,
           Percent = paste0(round((n / total) * 100, 0),"%")) %>%
    select(-total) %>%
    arrange(desc(n)) 
  
  names(df) <- c("", "Count", "Percent")
  pander(df, big.mark = ',')
}

str_left = function(text, num_char) {
  substr(text, 1, num_char)
}
 
str_right = function(text, num_char) {
  substr(text, nchar(text) - (num_char-1), nchar(text))
}

employer_xlsx <- get_newest("G:/2018 Projects/850-UWBEC ESPRI THRIVE/Data/Employer Data Pulls/", "\\.xlsx$")
osos_csv <- get_newest("G:/2018 Projects/850-UWBEC ESPRI THRIVE/Data/OSOS Data Pulls/", "\\.csv$")
wrap_around_xlsx <- get_newest("G:/2018 Projects/850-UWBEC ESPRI THRIVE/Data/Wrap Around Data Pulls/", "\\.xlsx$")
xwalk <- read_excel(get_newest("G:/2018 Projects/850-UWBEC ESPRI THRIVE/Data/Data Crosswalks/", "\\.xlsx$"))

was_xwalk <- read_excel(get_newest("G:/2018 Projects/850-UWBEC ESPRI THRIVE/Data/Data Crosswalks/", "\\.xlsx$"), "was_xwalk")

# OSOS Data

osos_df <- read.csv(osos_csv) %>%
  rename(OSOSID = OSOS.ID) %>%
  filter(OSOSID != "")

## Employer Feedback Data

employer_df <- read_excel(employer_xlsx, sheet = "person")

employer_df <- read_excel(employer_xlsx, sheet = "application") %>%
  rename(job.id = application.job_id,
         organization.id = application.organization_id,
         person.id = application.person_id) %>%
  merge(employer_df, all.y = TRUE)

employer_df <- read_excel(employer_xlsx, sheet = "job") %>%
  rename(company.id = job.company_id) %>%
  merge(employer_df, all.y = TRUE)

employer_df <- read_excel(employer_xlsx, sheet = "company") %>%
  merge(employer_df, all.y = TRUE)

employer_df <- read_excel(employer_xlsx, sheet = "organization") %>%
  merge(employer_df, all.y = TRUE) %>%
  filter(person.deleted != 1) %>%
  arrange(person.id)

wrap_around_df <- read_excel(wrap_around_xlsx, sheet = "Funds", skip = 8) %>%
  rename(`Funds ID` = ...1,
         Name = ...2,
         `Staff Name` = ...3,
         Partner = ...4) %>%
  filter(!is.na(Name)) %>%
  mutate(`Funds ID` = na.locf(`Funds ID`)) %>%
  merge(xwalk) %>%
  rename(funds_id = `Funds ID`)
```
## OSOS Data

There are <b>`r nrow(osos_df)`</b> Thrive participants on OSOS.


### Active or Inactive

```{r}
quick_tally(osos_df, "Status", "Active or Inactive")
```

### Cultural Barriers

```{r}
quick_tally(osos_df, "Cultural.Barriers", "Cultural Barriers")
```

### Disability Status

```{r}
quick_tally(osos_df, "Dis..Status", "Disability Status")
```

### Employment Status

```{r}
quick_tally(osos_df, "Empl.Status", "Employment Status")
```

### English Language Learner

```{r}
quick_tally(osos_df, "Eng.Lang.Learner", "English Language Learner")
```

### Veterans

```{r}
quick_tally(osos_df, "Vet", "Veterans")
```

### Gender

```{r}
quick_tally(osos_df, "Gender", "Active or Inactive")
```

### Education Level

```{r}
quick_tally(osos_df, "Edu..Level", "Active or Inactive")
```

### Underemployed

```{r}
quick_tally(osos_df, "Underemployed", "Active or Inactive")
```
<!--
### Employment Objective

```{r}
quick_tally(osos_df, "Empl..Objective", "Active or Inactive")
```

### Desired Title

```{r}
quick_tally(osos_df, "Desired.Title", "Active or Inactive")
```
-->
## Wrap Around Services

```{r}
temp <- wrap_around_df %>%
  select(contains("100")) %>%
  gather(Category, value) %>%
  na.omit() %>%
  merge(was_xwalk)

grand_total <- sum(temp$value)
```

There have been `r nrow(temp)` wrap around services provided to `r max(wrap_around_df$funds_id)` unique participants.  In total there has been `r dollar(grand_total)` spent in wrap around services.  There are `r nrow(wrap_around_df)` records.

### Service Category

```{r}
quick_tally(temp, "Wrap Around Service", "Wrap Around Service Category")
```

### Detailed Service Category

```{r}
quick_tally(temp, "Category", "Service Category")
```

### Total Support by Service Category as of `r str_left(str_right(wrap_around_xlsx, 15),10)`

```{r}
temp2 <- temp %>%
  group_by(`Wrap Around Service`) %>%
  summarise(`Total Support` = sum(value)) 

grand_total2 <- temp2 %>%
  summarise(`Total Support`  = sum(`Total Support` )) %>%
  mutate(`Wrap Around Service` = "Grand Total", Percent = "100%")

temp2%>%
  arrange(desc(`Total Support`)) %>%
  mutate(Percent = paste0(round(`Total Support` / grand_total * 100, 0), "%")) %>%
  bind_rows(grand_total2) %>%
  rename(` ` = `Wrap Around Service`) %>%
  mutate(`Total Support` = dollar(`Total Support`)) %>%
  pander()
```

### Total Support by Detailed Service Category
```{r}
grand_total2 <- grand_total2 %>%
  rename(Category = `Wrap Around Service`)

temp %>%
  group_by(Category) %>%
  summarise(`Total Support` = sum(value)) %>%
  arrange(desc(`Total Support`)) %>%
  mutate(Percent = paste0(round(`Total Support` / grand_total * 100, 0), "%")) %>%
  bind_rows(grand_total2) %>%
  rename(` ` = Category) %>%
  mutate(`Total Support` = dollar(`Total Support`)) %>%
  pander()
```


### Number of Wrap Around Services
```{r}
total <- max(wrap_around_df$funds_id)
temp <- wrap_around_df %>%
  group_by(Name) %>%
  tally() %>%
  rename(Services = n) %>%
  group_by(Services) %>%
  tally() %>%
  mutate(Services = as.factor(Services),
         Percent = round(n / total*100,0)) 
temp %>%
  ggplot(aes(x = Services, y = Percent, fill = Services)) +
  geom_col() +
  scale_fill_manual(values=c("#b30838", "#c1d2e1", "#00446a", "#feefc7", "#eaaa00", "#c6d6a7")) +
  theme(legend.position = "none")
```

```{r}
temp %>%
  select(Services, Percent) %>%
  mutate(Percent = paste0(Percent, "%")) %>%
  pander(big.mark = ',')
```
<!--
### Particpant Name

```{r}
quick_tally(wrap_around_df, "Name", "Particpant Name")
```
-->

### Partner Name

```{r}
quick_tally(wrap_around_df, "Partner", "Partner Name")
```

### Funds per Client

```{r}
wrap_around_df %>%
  select(`Funds per client`) %>%
  na.omit() %>%
  ggplot(aes(`Funds per client`)) +
  geom_histogram(bins = 50, fill="#b20838") +
  ylab("Participants") +
  theme(legend.position = "none")
```

## Employer Feedback Data

### General Counts

* Companies: `r read_excel(employer_xlsx, sheet = "company") %>% filter(company.deleted != 1) %>% nrow()`
* Jobs: `r read_excel(employer_xlsx, sheet = "job") %>% filter(job.deleted != 1) %>% nrow()`
* People: `r read_excel(employer_xlsx, sheet = "person") %>% filter(person.deleted != 1) %>% nrow()`
* Interviews: `r employer_df %>% select(application.date_referred) %>% na.omit() %>% nrow()`

### Interviews

#### Month Interviewed

```{r}
employer_df %>%
  select(application.date_referred) %>%
  na.omit() %>%
  mutate(interviewed = format(as.Date(application.date_referred),"%Y-%m")) %>%
  quick_tally(., "interviewed", "Date Interviewed")
```
<!--
#### Participant's Name

```{r}
employer_df %>%
  select(person.full_name, job.description) %>%
  na.omit() %>%
  quick_tally(., "person.full_name", "Participant's Name")
```
-->
#### Partner Name

```{r}
employer_df %>%
  select(organization.name) %>%
  na.omit() %>%
  quick_tally(., "organization.name", "Partner Name")
```

### Hires

#### Date Hired

```{r}
temp <- employer_df %>%
  select(application.date_hired) %>%
  na.omit() %>%
  mutate(hired = format(as.Date(application.date_hired),"%Y-%m")) 
if(nrow(temp) > 0) {
  temp %>%
    quick_tally(., "hired", "Date Hired")
} else {
  cat("No data from partners")
}
```
<!--
#### Participant's Name

```{r}
employer_df %>%
  select(person.full_name, application.date_hired) %>%
  na.omit() %>%
  quick_tally(., "person.full_name", "Participant's Name")
```
-->
### Fully Productive

```{r}
temp <- employer_df %>%
  select(application.date_fully_productive) %>%
  na.omit() %>%
  mutate(fully_productive = format(as.Date(application.date_fully_productive),"%Y-%m")) 
if(nrow(temp) > 0) {
  temp %>%
    quick_tally(., "fully_productive", "Fully Productive")
} else {
  cat("No data from partners")
}
```

<!--
#### Participant's Name

```{r}
employer_df %>%
  select(person.full_name, application.date_fully_productive) %>%
  na.omit() %>%
  quick_tally(., "person.full_name", "Participant's Name")
```
-->
*Source Data Files*

* `r employer_xlsx`
* `r osos_csv`
* `r wrap_around_xlsx`
